disaggregation: An \proglang{R} package for Bayesian spatial disaggregation modelling

### 2019-12-02 article.Rnw
Adding Rohan's edits

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

@@

Suppose we have response data, $y_j$, for $N$ polygons, which corresponds to count data for the property of interest within that polygon. 
The process that is being counted occurs in continuous space that we model as a high-resolution, square lattice for convenience.
The data, $y_j$,  are assumed to be created by the aggregation of the counts over the polygon, i.e. the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. 
The rate, $p_i$, is defined such that the number of cases in a pixel can be calculated by multiplying the rate by the aggregation raster. For example, in epidemiology, we may have as our response the number of people that contract a certain disease in a given period of time (case incidence). Our rate would be the incidence rate of cases per population, where the aggregation raster corresponds to population.

For the disaggregation model, we first model the rate, $pred_i$, at pixel level:
%
\begin{equation} \label{eq:lp}
link(pred_i) = \beta_0 + \beta X + GP(s_i) + u_j
\end{equation}
%

where $\beta$ are the regression coefficients, $X$ are the covariates, $GP$ is the gaussian random field and $u_j$ is the polygon-specific iid random effect. The user defined link function is given by $link$. The Gaussian random field has a Matern covariance function, with hyperparameters $\rho = \frac{\sqrt{8\lambda}}{\kappa}$, the range of the field (beyond which correlation is $<$ 0.1), and $\sigma$, the marginal standard deviation, relating the magntiude of the variation of the field. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

The pixel predictions are then aggregated to the polygon level using the weighted sum (via the aggregation raster, $agg_i$):

%
\begin{equation} \label{eq:agreggate1}
cases_j = \sum_{i \epsilon j} pred_i \times agg_i
\end{equation}
\begin{equation} \label{eq:agreggate2}
rate_j = \frac{\sum_{i \epsilon j} pred_i \times agg_i}{\sum_{i \epsilon j} agg_i}
\end{equation}
%

The different likelihoods correspond to slightly different models ($y_j$ is the response count data):

%
\begin{itemize}
\item {\textbf{Gaussian} ($\sigma_j$ is the dispersion of the polygon data)
\begin{equation} \label{eq:gaussian}
y_j \sim Normal(cases_j/\sum agg_i, rate_j, \sigma_j)
\end{equation}
Here $\sigma_j = \sigma \sqrt{\sum agg_i^2} / \sum agg_i$, where  $\sigma$ is the dispersion of the pixel data (a parameter learnt by the model) }

\item \textbf{Binomial} (For a survey in polygon $j$, $y_j$ is the number positive cases and $N_j$ is the number tested),
\begin{equation} \label{eq:binomial}
y_j \sim Binom(cases_j, N_j, rate_j)
\end{equation}

\item \textbf{Poisson} (predicts incidence count),
\begin{equation} \label{eq:poisson}
y_j \sim Pois(cases_j)
\end{equation}

@@

For each of the model parameters and hyperparameters we specify priors and hyperpriors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$.

The field parameters are given penalised complexity priors \cite{fuglstad2018constructing}, $\rho_{min}$ and $\rho_{prob}$, where $P(\rho < \rho_{min}) = \rho_{prob}$, and the variation, $\sigma_{max}$ and $\sigma_{prob}$, where $P(\sigma > \sigma_{max}) = \sigma_{prob}$. The effect of these hyperpriors is to regularise the field to have smaller variations over larger distances. Similarly for the iid effect, the priors are penalised complexity priors \cite{simpson2017penalising} on the standard deviation, $sd_{max}$ and $sd_{prob}$, such that $P(sd > sd_{max}) = sd_{prob}$. This has the effect of regularising the iid effect to be small.

@@

In this section we show performance comparisons between modelling using the Laplace approximation provided by the \pkg{disaggregation} package, based on \pkg{TMB}, and using Markov chain Monte Carlo (MCMC). MCMC algorithms sample the probability space to produce an esimate of the probability distribution. These algorithms tend to be slow, as it takes a long time to effectively sample the likelihood space. In contrast, the \pkg{disaggregation} package, which uses the Laplace approximation, automatic differentiation and efficient computation of sparse matrices within \pkg{TMB}, takes a considerably shorter time to optimise the model. Here we compare the time and performance of the two techniques.

>>>>

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\usepackage{amsmath}

@@

Suppose we have response data, $y_i$, for $N$ polygons, which corresponds to count data for the property of interest within that polygon. 
The process that is being counted occurs in continuous space that we model as a high-resolution, square lattice for convenience.
The data, $y_i$,  are assumed to be created by the aggregation of the counts over the polygon, i.e. the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. 
The rate is defined such that the number of cases in this pixel can be calculated by multiplying the rate by the aggregation raster. For example, in epidemiology, we may have as our response the number of people that contract a certain disease in a given period of time (case incidence). Our rate would be the incidence rate of cases per population, where the aggregation raster corresponds to population.

For the disaggregation model, we first model the rate at pixel level, with the rate in pixel $j$ of polygon $i$ given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
Here $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

The pixel predictions are then aggregated to the polygon level using the weighted sum (via the aggregation raster $a_{ij}$):

%
\begin{equation} \label{eq:agreggate1}
\textrm{cases}_i = \sum_{j=1}^{N_i} a_{ij}\textrm{rate}_{ij}
\end{equation}
\begin{equation} \label{eq:agreggate2}
\textrm{rate}_i = \frac{\textrm{cases}_{i}}{\sum_{j=1}^{N_i} a_{ij}}
\end{equation}
%
The different likelihoods correspond to slightly different models ($y_i$ is the response count data):

%
\begin{itemize}
\item {\textbf{Gaussian} ($\sigma_i$ is the dispersion of the polygon data)
\begin{equation} \label{eq:gaussian}
y_i \sim \textrm{Normal}(\textrm{cases}_i, \sigma_i)
\end{equation}
Here $\sigma_i = \sigma \sqrt{\sum_j a_{ij}^2}$, where  $\sigma$ is the pixel-level dispersion (a parameter learnt by the model) }

\item \textbf{Binomial} (For a survey in polygon $i$, $y_i$ is the number positive cases and $M_i$ is the number tested),
\begin{equation} \label{eq:binomial}
y_i \sim \textrm{Binomial}(M_i, \textrm{rate}_i)
\end{equation}

\item \textbf{Poisson} (predicts incidence count),
\begin{equation} \label{eq:poisson}
y_i \sim \textrm{Pois}(\textrm{cases}_i)
\end{equation}

@@

For each of the model parameters and hyperparameters we specify priors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$. A penalised complexity prior is placed on the scale and range parameters of the random field \cite{fuglstad2018constructing} such that
\begin{align}
    P(\rho < \rho_{min}) &= \rho_{prob} \\
    P(\sigma > \sigma_{max}) &= \sigma_{prob}
\end{align}
where the values $\rho_{min}, \,\rho_{prob}, \,\sigma_{max}, \,\sigma_{prob}$ are set by the user. This prior shrinks the field towards a base model with zero variance and infinite range, in other words regularising towards a flatter field with smaller magnitude. The polygon-specific effects $u_1,...,u_N$ have Gaussian priors centered at 0 with standard deviation $u_{sd}$. A penalised complexity prior \cite{simpson2017penalising} such that $P(u_{sd} > sd_{max}) = sd_{prob}$ which shrinks towards a base model of no polygon-specific effect. The $sd_{max}$ and $sd_{prob}$ are set by the user.

@@

In this section we show performance comparisons between modelling using the Laplace approximation provided by the \pkg{disaggregation} package, based on \pkg{TMB}, and using Markov chain Monte Carlo (MCMC). Given a function to evaluate the probability density of a distribution at any given point in parameter space, MCMC algorithms construct markov chains to generate samples from this distribution. These algorithms are often slow, particularly in high-dimensional settings, as it can take a long time to converge to and effectively sample from the stationary distribution. The density is evaluated at each step, so this problem is compounded when evaluating this density is computationally expensive. In contrast, the \pkg{disaggregation} package approximates uses a Laplace approximation to the posterior to generate posterior samples. This only requires the posterior to be maximised to find the posterior mode and therefore involves relatively few evaluations of the posterior density compared to MCMC techniques, although potentially at the expense of less accurate posterior samples. Here we compare the time and performance of the two techniques.
